Nail: A practical interface generator for data formats

### 2014-04-02 paper/design.tex
Design

A central design decision of Nail is that there is a semantic bijection between the model exposed to
the programmer and the bytes input and output up to syntactic equivalence. More precisely, the
parser is the generators inverse, so parsing the generators output will yield the generators input,
but generating the parsers output does not necessarily yield the parsers input. To understand why
this makes sense, consider a grammar for a text language that tolerates white space or a binary
protocol that tolerates arbitrarily long padding\footnote{Say, the physical layer of most
  communication protocols is a possibly infinite sequence of symbols that are syntactically nil
  followed by a pre-determined synchronisation sequence and the actual contents of the
  transmission.}. In that case, the program semantics should be independent of the number of padding
elements in the input, and Nail therefore does not expose that information to the programmer. We
call such discarded fields \emph{constants}. Currently, Nail always makes a default choice when
there are multiple options to express a constant, however Nail could be extended to allow a
grammar-specific plug-in to make these choices, say for faster alignment, streaming applications
when data is not ready or visual appearance in human-readable protocols. However, if some features
of Nail are avoided, there is a proper isomorphism between model and protocol.

Nail evaluates the combinators and produces type declarations for the internal model, the \textit{parser}, a function to parse a sequence of
bytes into an instance of the model and the \textit{generator}, a function to create a sequence of
bytes from an instance of the model.

In the following, we present the combinators in Nail.
\paragraph{Fundamental parsers}
The elementary parsers of Nail are the same as those of Hammer, signed and unsigned integers with
arbitrary lengths up to 64 bits. Note that is possible to define parsers for sub-byte lengths, e.g.
to parse the 4-bit data offset within the TCP header. Integer parsers return their value in the
smallest appropriately sized machine integer type, e.g. a 24 bit integer is stored in a 32-bit wide variable.\\
Examples: \texttt{uint4}, \texttt{int32}

Integer parsers can be constrained to fall either within an (inclusive) range of values or be an
element of a set of allowed values. \\
Examples: \texttt{uint8 | 1..3} (accepts an 8 bit unsigned
integer between 1 and 3 inclusive) ,\\
 \texttt{uint16| ..512} (an unsigned 16-bit integer less or
equal than 512),\\
\texttt{int32| [1,255,512]} (a signed 32 bit integer with value 1, 255 or 512).

\paragraph{Constant parser}
Nail also has constant parsers, which do not return a value. The simplest constant parser is an
integer with fixed value, e.g. \texttt{uint8=0}. Constant parsers succeed if the inner integer
parser returns the given value and fail otherwise.

\paragraph{Sequence: Structure}
Nails fundamental concept is the structure combinator. It contains a list of
named parsers and unnamed constant parsers. When parsing, it parses each parser
in sequence and returns a structure containing the result of each parser.

\begin{tabular}{ll}
\textbf{Nail} & \textbf{Data Model}\\
\begin{minipage}{2in}
\begin{verbatim}
header = {
 id uint16
 qr uint1
 opcode uint4
 aa uint1 
 tc uint1
 rd uint1
 ra uint1
 uint3 = 0
 rcode uint4
}
\end{verbatim} 
\end{minipage}
 & 
\begin{minipage}{2in}
\begin{verbatim}
struct header{
    uint16_t id;
    uint8_t qr;
    uint8_t opcode;
    uint8_t aa;
    uint8_t tc;
    uint8_t rd;
    uint8_t ra;
    uint8_t rcode;
};
\end{verbatim} 
\end{minipage} 
\\
\end{tabular}
\paragraph{Repetition: Many,SepBy}
The many combinator takes a parser and applies it repeatedly until it fails, returning an array of
the inner parsers results. SepBy additionally takes a constant parser, which it applies in between
parsing two values,but not before parsing the first value or after parsing the last.\\
Examples \texttt{many uint8}, \texttt{sepBy uint8=',' (many uint8| '0'..'9')} (which recognises
comma-separated lists of numbers).

The many combinator also works on constant parsers, although as described above it can make the
generator not bijective.\footnote{By default, the generator will emit exactly one iteration of the
  constant parser in its output.} Example: \texttt{many uint8= "test"}

\paragraph{Semantic choice: choose}
We extend PEG's ordered choice combinator, which picks the first choice that succeeds, with a tag
for each choice. The result is saved in a tagged union.

\begin{verbatim}
choose {
   A = uint8 | 1.. 8
   B = uint16 | ..256
}
\end{verbatim}
\paragraph{Wrap}

The wrap combinator parses one or more constant parser, then an inner parser whose value it returns
and then again parses one or more constant parsers.\\ Example: 
\verb+<uint8='"'; many8 uint8|'a'..'z'; uint8='"'  >+

\subsection{Offset fields}
Another problem for parser generators is that binary protocols often contain length and offset
fields. While conventional parsing algorithms can deal with bounded offset fields - a finite
automaton can count a bounded integer, and we can feed the (finite) input multiple times to the
finite automaton. However, this implementation is both time-inefficient - why feed many bytes into
the automaton that will just be skipped - and very cumbersome to express with current parser
generators. Therefore, if languages with offset fields are parsed with parser generators, the only
currently feasible way is to add 'ad-hoc' hacks such as changing the input pointer of the generated
parser on the fly from semantic actions.

@@

\paragraph{Dependency fields}
We call length or offset fields \textit{dependency fields}, because during parsing, another parser
depends on them, and while generating output, their value depends on some other structure in the
data model.
Dependency fields appear in a structure combinator as would any other field, but their name begins
with an '@' sign and they have to return an integer value. Dependency fields are not directly
represented in the internal model.

\paragraph{Length fields: n_of}
The length combinator takes a dependency field and a parser, evaluating the parser as often as the
value of the dependency field, returning an array of the values. When generating output, it emits
the array and writes its length to the dependency field.

\subsection{Example Grammars}
To further familiarise the reader with Nail, we provide two example grammars. 
\paragraph{UTF-16}
This grammar recognises valid UTF-16 strings and exposes an array of code points.

>>>>

A central design decision of Nail is that there is a semantic bijection between
the model exposed to the programmer and the bytes input and output up to
syntactic equivalence for unambiguous grammars. More precisely, the parser is
the generators inverse, so parsing the generators output will yield the
generators input, but generating the parsers output does not necessarily yield
the parsers input. To understand why this makes sense, consider a grammar for a
text language that tolerates white space or a binary protocol that tolerates
arbitrarily long padding\footnote{Say, the physical layer of most communication
  protocols is a possibly infinite sequence of symbols that are syntactically
  nil followed by a pre-determined synchronisation sequence and the actual
  contents of the transmission.}. In that case, the program semantics should be
independent of the number of padding elements in the input, and Nail therefore
does not expose that information to the programmer. We call such discarded
fields \emph{constants}. Currently, Nail always makes a default choice when
there are multiple options to express a constant, however Nail could be extended
to allow a grammar-specific plug-in to make these choices, say for faster
alignment, streaming applications when data is not ready or visual appearance in
human-readable protocols. Similarly, Nail does not preserve the layout of
objects referred to by their offsets. If the grammar contains no constants and
offset fields, there is a proper isomorphism between model and protocol.

Nail evaluates the combinators and produces type declarations for the internal
model, the \textit{parser}, a function to parse a sequence of bytes into an
instance of the model and the \textit{generator}, a function to create a
sequence of bytes from an instance of the model. In the following, we present
Nail's combinators.

\paragraph{Fundamental parsers}
The elementary parsers of Nail are the same as those of Hammer, signed and
unsigned integers with arbitrary lengths up to 64 bits. Note that is possible to
define parsers for sub-byte lengths, e.g. to parse the 4-bit data offset within
the TCP header. Integer parsers return their value in the
smallest appropriately sized machine integer type, e.g. a 24 bit integer is stored in a 32-bit wide variable.\\
Examples: \texttt{uint4}, \texttt{int32}

Integer parsers can be constrained to fall either within an (inclusive) range of
values or be an
element of a set of allowed values. \\
Examples:\\
\texttt{uint8 | 1..3} (8 bit unsigned integer $1\leq x \leq 3$) ,\\
\texttt{uint16| ..512} (unsigned 16-bit integer $\leq 512$),\\
\texttt{int32 | [1,255,512]} (signed 32 bit integer with value 1, 255 or 512).
Invalid values or prematurely reaching the end of input raise an error when
parsing input or generating output and abort the procedure.

\paragraph{Sequence: Structure}
Nails fundamental concept is the structure combinator. It contains a list of
named parsers and unnamed constant parsers  . The parser invokes each field 
in sequence and returns a structure containing the result of each parser.

\begin{tabular}{ll}
  \textbf{Nail} & \textbf{Data Model}\\
  \begin{minipage}{1.2in}
\begin{verbatim}
header = {
 id uint16
 qr uint1
 opcode uint4
 aa uint1 
 tc uint1
 rd uint1
 ra uint1
 uint3 = 0
 rcode uint4
}
\end{verbatim} 
  \end{minipage}
  & 
  \begin{minipage}{1.2in}
\begin{verbatim}
struct header{
    uint16_t id;
    uint8_t qr;
    uint8_t opcode;
    uint8_t aa;
    uint8_t tc;
    uint8_t rd;
    uint8_t ra;
    uint8_t rcode;
};
\end{verbatim} 
  \end{minipage}
  \\
\end{tabular}
\paragraph{Repetition: Many,SepBy}
The many combinator takes a parser and applies it repeatedly until it fails,
returning an array of the inner parsers results. SepBy additionally takes a
constant parser, which it applies in between
parsing two values,but not before parsing the first value or after parsing the last.\\
Examples: \texttt{many uint8}\\ \texttt{sepBy uint8=',' (many uint8| '0'..'9')}
(which recognises comma-separated lists of numbers).

\paragraph{Semantic choice: choose}
We extend PEG's ordered choice combinator with a tag for each choice. The parser
attempts to parse each option in the order they are specified in the field and
stores the result in a tagged union. If an option fails, the parser backtracks
to the beginning of the choice combinator's input. Care must be taken that the
choices do not overlap, because Nail always picks to the first successfully
parsed choice. If two options overlap, generated output for the latter option is
not necessarily understood identically by the parser. However, actual data
formats are usually not ambiguous in this sense.

\begin{verbatim}
choose {
   A = uint8 | 1.. 8
   B = uint16 | ..256
}
\end{verbatim}

\subsection{Constant parsers}
Nail also features \emph{constant parsers}, which are not affecting the internal
representation. Constant parsers can appear instead of structure fields and as
separators in the SepBy combinator.

The simplest constant parser is an integer or array of integers with fixed
value, e.g. \texttt{uint8=0}, \texttt{many uint8=[1,2]}. A convenience notation
for strings is also supported: \texttt{many uint8 = "foo"}.

\paragraph{Wrap combinator.} When implementing real protocols with Nail, we often found that many structures
consist of many constant parsers and only one named field. This pattern is
common in binary protocols which use fixed headers to denote the type of data
structure to be parsed.  In order to keep the internal representation cleaner,
we introduced the wrap combinator, which takes a sequence of parsers containing
exactly one non-constant parser. The parser and generator act as though the wrap
combinator was a sequence of parsers, but the data model does not wrap the
single value in another structure, making the application code more concise.\\
Example: \\ \verb+<uint8='"';many int8|'a'..'z'; uint8='"'>+  parses a quoted
lowercase word into an array - excluding the quotation marks.

\paragraph{Constant combinators.}
In some protocols, there might be many ways to represent the same constant field
and there is no semantic difference between different syntaxes.  Nail therefore
supports choice (\texttt{|| uint8 = 0 || uint16 = 0xFFFF}) and repetition
(\texttt{many (uint8= ' ')}). This is particularly useful for handling padding or whitespace.

As discussed above, choosing to use these combinators on constant parsers
removes the bijection between byte-strings and our data model, as there are
multiple byte-strings that correspond to the same internal data structure and
the generator has to choose one of these representations. As such, constant
combinators are the generator dual of ambiguous choice combinators in the
parser, because they lead to ambiguities in the generator.

\subsection{Dependent fields}
Another problem for parser generators is that binary protocols often contain
length and offset fields. While conventional parsing algorithms can deal with
bounded offset fields - a finite automaton can count a bounded integer, and we
can feed the (finite) input multiple times to the finite automaton. However,
this implementation is both time-inefficient - why feed many bytes into the
automaton that will just be skipped - and very cumbersome to express with
current parser generators. Therefore, if languages with offset fields are parsed
with parser generators, the only currently feasible way is to add 'ad-hoc' hacks
such as changing the input pointer of the generated parser on the fly from
semantic actions.

@@

We call length or offset fields \textit{dependent fields}, because during
parsing, another parser depends on them, and while generating output, their
value depends on some other structure in the data model. Dependency fields
appear in a structure combinator as would any other integer field, but their
name begins with an '@' sign. Dependency fields are not exposed in the data
model, but transparently computed.

A dependency field has to appear in the grammar before it can be used. 
\paragraph{Length fields: n_of}
The length combinator takes a dependency field and a parser, evaluating the
parser as often as the value of the dependency field, returning an array of the
values. When generating output, it emits the array and writes its length to the
dependency field.
\paragraph{Offsets: offset} 
The offset combinator takes a dependency field and a parser. It moves the parser
to the position specified in the offset field and invokes the inner parser, and
then moves the input back to its original position.
While generating output, all structures referred to by offset are generated
after the main structure and the dependent offset fields patched up.

\paragraph{Checksums: raw_depend}
In many data formats, some values depend on external representation, such as
checksums and cryptographic signatures. While it would be possible to extend
our constraint language to be powerful enough to support such constructs, we
would essentially be building a separate, Turing complete language that has all
the same pitfalls existing programs have. Therefore, we intend to  allow the programmer to
very carefully escape Nail's programming model and write a function that takes a
pointer to the dependent value and a range of bytes. 
This feature is not implemented in the current prototype. \\
Example:\\
 \verb+{ data many uint8; @checksum uint32+\\
 \verb+raw_depend @checksum data crc32 }+\\
 \verb+/*bool crc32(uint32_t *out, uint8_t *in);*/+
Because this feature compromises Nail's security guarantees, it should only be
used in limited circumstances and with carefully prepared checksum functions. 
\subsection{Example Grammars}
To further familiarise the reader with Nail, we provide two example grammars.
\paragraph{UTF-16}
This grammar recognises valid UTF-16 strings and exposes an array of code
points.
